\section{Background and the MC-PILOT Algorithm}
\label{sec:background}

\subsection{Problem Formulation}
\label{sec:background:problem}

\paragraph{MDP and state space.}
Robotic throwing is modelled as a single-shot Markov decision process.
The ball state $\mathbf{x} = [\mathbf{p},\, \dot{\mathbf{p}}] \in \mathbb{R}^6$
concatenates 3D position and velocity.
Because the policy must generalise across a 2D target domain, the state is augmented
with the target position $\mathbf{P} = (P_x, P_y) \in \mathbb{R}^2$ to form
\begin{equation}
  \tilde{\mathbf{x}} = [\mathbf{x},\, \mathbf{P}] \in \mathbb{R}^8.
  \label{eq:augmented_state}
\end{equation}
The GP learns \emph{geometry-conditioned} residual corrections in this augmented space,
enabling a single model to serve all targets without per-target retraining.

\paragraph{Landing cost.}
Performance is measured at the terminal timestep $T$ (ball landing) by a
Gaussian-shaped cost:
\begin{equation}
  c(\tilde{\mathbf{x}}_T) = 1 - \exp\!\left(
    -\frac{\|\mathbf{p}_T - \mathbf{P}\|^2}{\ell_c^2}
  \right),
  \label{eq:cost}
\end{equation}
where $\mathbf{p}_T \in \mathbb{R}^2$ is the landing position and $\ell_c$ controls
the acceptance width.
The policy objective is $J(\theta) = \mathbb{E}[c(\tilde{\mathbf{x}}_T)]$, minimised
by gradient descent through Monte Carlo rollouts.

\subsection{Gaussian Process Dynamics}
\label{sec:background:gp}

The GP maintains a posterior over per-timestep velocity corrections
$[\Delta v_x, \Delta v_y, \Delta v_z]$ conditioned on the augmented state
$\tilde{\mathbf{x}}_t$.
Let $\mathcal{D} = \{(\tilde{\mathbf{x}}^{(k)},\, \Delta \mathbf{v}^{(k)})\}_{k=1}^N$
denote the observed transitions.
A squared-exponential (SE) kernel
$k(\tilde{\mathbf{x}},\, \tilde{\mathbf{x}}') = \sigma_f^2 \exp(-\|\tilde{\mathbf{x}}
- \tilde{\mathbf{x}}'\|^2 / 2\ell^2)$
captures smooth velocity-correction structure.
The posterior mean provides the best single-point dynamics prediction; the
posterior variance $\sigma^2(\tilde{\mathbf{x}})$ quantifies epistemic uncertainty and
drives the optimisation to avoid regions of high model uncertainty~\cite{rasmussen2006gp}.

\paragraph{Why GPs?}
Three properties make GP dynamics especially appropriate here.
First, \emph{exact uncertainty quantification}: the posterior variance is meaningful
even with five training points, so policy gradients know how confident the model is at
every state.
Second, \emph{data efficiency}: the kernel encodes smoothness priors that allow
generalisation from as few as five exploration throws.
Third, \emph{Bayesian posterior updates}: each throw updates the posterior analytically,
yielding a dynamics model that improves monotonically with data.

\paragraph{Sparse approximation.}
Exact GP inference scales as $O(N^3)$ in the number of observations.
As data accumulate across trials (5 exploration + up to 10 learning throws, each
producing $\approx 35$ state transitions), this becomes the dominant computational cost.
\mcpilco{} addresses this with a Subset-of-Data (SOD) approximation: a representative
inducing set of $M_{\mathrm{ind}}$ points is selected from the full training set, and
inference is performed using only these inducing points.
The key guarantee is that inducing set selection does not discard the most informative
transitions, so GP posterior quality degrades gracefully rather than catastrophically as
the dataset grows.
In our experiments, a budget of $M_{\mathrm{ind}} = 100$ inducing points was sufficient
to maintain accurate dynamics predictions throughout all ten learning trials.

\paragraph{Speed-integration model.}
The state update follows the reconstruction rule of \cite{turcato2025mcpilot} (Eq.~18):
\begin{align}
  \mathbf{v}_{t+1} &= \mathbf{v}_t + \Delta\mathbf{v}_t, \label{eq:vel_update} \\
  \mathbf{p}_{t+1} &= \mathbf{p}_t + T_s\,\mathbf{v}_t + \tfrac{T_s}{2}\,\Delta\mathbf{v}_t,
  \label{eq:pos_update}
\end{align}
where $T_s = 0.02$\,s is the integration timestep and $\Delta\mathbf{v}_t$ is the GP
residual correction (incorporating gravity and drag).
This midpoint rule reduces the systematic position error relative to a forward-Euler
update at the same timestep.

\subsection{Monte Carlo Policy Gradient}
\label{sec:background:mcpg}

\mcpilco{}~\cite{mcpilco} propagates $M$ stochastic particles through the GP posterior
using the \emph{reparameterisation trick}~\cite{kingma2014auto}.
Rather than sampling GP function values directly (which would break the computation
graph), each GP draw is written as a deterministic transformation of a base noise
variable: $\mathbf{f}(\tilde{\mathbf{x}}) = \boldsymbol{\mu}(\tilde{\mathbf{x}}) +
L(\tilde{\mathbf{x}})\,\boldsymbol{\varepsilon}^{(m)}$, where $\boldsymbol{\mu}$ is
the GP posterior mean, $L$ is the Cholesky factor of the posterior covariance, and
$\boldsymbol{\varepsilon}^{(m)} \sim \mathcal{N}(\mathbf{0}, I)$ is sampled once per
particle.
Because $\boldsymbol{\varepsilon}^{(m)}$ is independent of $\theta$ and the
transformation through $L$ and $\boldsymbol{\mu}$ is differentiable, gradients flow
from the cost through the full rollout back to the policy parameters.
Each particle integrates $\tilde{\mathbf{x}}_{t+1}^{(m)} = g(\tilde{\mathbf{x}}_t^{(m)},\,
\pi_\theta(\tilde{\mathbf{x}}_t^{(m)})) + \mathbf{f}(\tilde{\mathbf{x}}_t^{(m)})$
for $T$ steps, yielding a set of differentiable rollouts
$\{\tilde{\mathbf{x}}_t^{(m)}\}_{t=0}^T$.
The policy gradient is approximated by
\begin{equation}
  \nabla_\theta J(\theta) \approx \frac{1}{M}\sum_{m=1}^{M} \nabla_\theta
    c\!\left(\tilde{\mathbf{x}}_T^{(m)}\right),
  \label{eq:mcpilco_gradient}
\end{equation}
and optimised with Adam~\cite{kingma2015adam} for $N_{\mathrm{opt}} = 1500$ steps per
trial.

\paragraph{Why Monte Carlo over analytic moment-matching?}
PILCO~\cite{pilco} propagates the full state distribution analytically by matching
Gaussian moments through the GP.
This is exact for Gaussian initial distributions but requires a Gaussian-closure
approximation at every step, which accumulates errors for multi-step rollouts.
More critically for the throwing problem, it breaks when the initial state is
\emph{non-Gaussian}.
The throwing policy $\pi_\theta(\mathbf{P}) \to v$ is a nonlinear function
(Eq.~\ref{eq:policy}) of the target position $\mathbf{P}$, which is sampled uniformly
across the workspace.
The resulting initial velocity $\mathbf{v} = (v\cos\alpha\cos\phi,\,
v\cos\alpha\sin\phi,\, v\sin\alpha)$ is a nonlinear function of $\mathbf{P}$, and
its distribution is non-Gaussian.
PILCO's moment-matching closure would introduce a systematic bias at $t=0$; the Monte
Carlo approach propagates the true empirical distribution without any distributional
assumption.
The variance of the MC gradient estimate decreases as $O(1/\sqrt{M})$ with particle
count $M = 400$.

\subsection{The Squashed-RBF Policy}
\label{sec:background:policy}

The throwing policy maps target position to release speed via a radial basis function
network with a $\tanh$ squashing output:
\begin{equation}
  \pi_\theta(\mathbf{P}) = \frac{u_M}{2}\!\left(
    \tanh\!\left(
      \sum_{i=1}^{N_b} w_i \exp\!\left(
        -\frac{\|\mathbf{a}_i - \mathbf{P}\|^2}{\ell_s^2}
      \right)
    \right) + 1
  \right),
  \label{eq:policy}
\end{equation}
where $u_M$ is the maximum allowable speed, $\{(\mathbf{a}_i, w_i)\}_{i=1}^{N_b}$ are
the $N_b$ RBF centres and weights (the trainable parameters $\theta$), and $\ell_s$ is
the spatial lengthscale controlling the influence radius of each basis function.
The $\tanh(\cdot) + 1$ construction squashes the weighted sum to $(0,\, u_M)$, enforcing
physical speed limits without box constraints on $\theta$.

\paragraph{Geometric interpretation.}
The $N_b = 250$ RBF centres tile the 2D target workspace uniformly.
Each centre $\mathbf{a}_i$ acts as a local expert for targets near it; the lengthscale
$\ell_s$ governs how far that expertise extends.
When $\ell_s$ is commensurate with the target range, different centres fire for
different targets and the policy can assign distinct speeds; when $\ell_s \gg
\text{target range}$, all centres fire together and the policy collapses to a constant
(Section~\ref{sec:convergence:lengthscale}).

\subsection{MC-PILOT Specific Modifications}
\label{sec:background:mcpilot}

\mcpilot{}~\cite{turcato2025mcpilot} adapts \mcpilco{} for throwing with two targeted
modifications over the standard continuous-control formulation.

\paragraph{Single-shot policy.}
Standard \mcpilco{} fires the policy at every timestep, which is appropriate for
balancing tasks requiring continuous feedback.
Throwing is fundamentally different: the projectile is in free flight after release, so
feedback control is neither possible nor necessary.
The policy collapses to a single map $\pi_\theta(\mathbf{P}) \to v$, called once at
$t{=}0$.
This dramatically reduces the parameter search space, explains the extreme data
efficiency, and is what allows the algorithm to converge in 5--10 trials.

\paragraph{Gripper delay estimation.}
A real gripper does not open instantaneously; the random delay $t_d \sim \mathcal{U}(a,b)$
causes the arm to decelerate for $t_d$ seconds past the commanded release instant,
introducing a systematic undershoot $\mathbf{v}_{\mathrm{actual}} \approx
(1 - \gamma t_d / \|\mathbf{v}_{\mathrm{cmd}}\|)\,\mathbf{v}_{\mathrm{cmd}}$.
\mcpilot{} estimates $(a, b)$ via Bayesian Optimisation~\cite{shahriari2016bo} on
hardware calibration throws, making the algorithm deployable on a real arm.
We replace this with injected stochastic noise in our simulation study
(Section~\ref{sec:study3}), which lets us vary noise type and magnitude in a controlled
manner while preserving GP-gradient compatibility.

\subsection{Aerodynamic Drag}
\label{sec:background:drag}

Ball trajectories include aerodynamic drag following Eq.~35 of~\cite{turcato2025mcpilot}:
\begin{equation}
  \mathbf{F}_D = -\tfrac{1}{2}\rho\, C_D(\mathrm{Re})\, A\,
    \|\dot{\mathbf{p}}\|\, \dot{\mathbf{p}},
  \label{eq:drag}
\end{equation}
where $\rho$ is air density, $A = \pi r^2$ is the ball cross-section, and $C_D$ depends
on Reynolds number $\mathrm{Re} = \|\dot{\mathbf{p}}\| \cdot 2r / \nu$.
For the tennis ball ($m = 0.0577$\,kg, $r = 0.0327$\,m) and speeds below 4\,m/s, $\mathrm{Re}$
is in the subcritical regime and $C_D \approx 0.47$ is constant.
Under wind $\mathbf{w}$, the physically correct formulation substitutes
$\mathbf{v}_{\mathrm{rel}} = \dot{\mathbf{p}} - \mathbf{w}$ for $\dot{\mathbf{p}}$
everywhere in Equation~\eqref{eq:drag}; this correction is implemented in
Study~4 (Section~\ref{sec:study4}).

\subsection{Algorithmic Lineage}
\label{sec:background:lineage}

Table~\ref{tab:lineage} places \mcpilot{} in the context of prior model-based RL for
continuous control, highlighting the key idea and data budget at each step.

\begin{table}[t]
  \centering
  \caption{Algorithmic lineage from model-free RL to \mcpilot{}.}
  \label{tab:lineage}
  \small
  \begin{tabular}{llcc}
    \toprule
    Algorithm & Key Idea & Data Needed & Year \\
    \midrule
    Model-free RL & Learn policy from reward only   & $10^3$--$10^4$ & 2013+ \\
    TossingBot~\cite{tossbot} & Residual physics + DNN & $\sim$500    & 2020 \\
    PILCO~\cite{pilco}        & GP dynamics + analytic gradient & $\sim$50 & 2011 \\
    \mcpilco~\cite{mcpilco}   & MC gradient through GP          & $\sim$20--50 & 2022 \\
    \textbf{\mcpilot}~\cite{turcato2025mcpilot} & Single-shot + gripper BO & \textbf{$\sim$10} & 2025 \\
    \bottomrule
  \end{tabular}
\end{table}
