\section{Conclusion}
\label{sec:conclusion}

We presented a diagnostic reproduction of \mcpilot{}~\cite{turcato2025mcpilot} in a
self-contained NumPy and PyBullet simulator and conducted five experimental studies to
characterise the algorithm's robustness, portability, and limits.

\paragraph{Implementation.}
Four non-obvious silent failure modes were diagnosed and corrected: a broken gradient
path from cost to policy parameters, post-landing GP hallucination from particles
propagating underground, cost-lengthscale saturation that kills gradients during
exploration, and a simulation-horizon unit mismatch that exhausted RAM.
Each failure was identified by instrumentation, understood through the algorithm
structure, and resolved with a minimal targeted fix.

\paragraph{Convergence rules.}
Two transferable design rules emerged.
The \emph{lengthscale-to-range scaling rule} $\ell_s \approx 0.15 \times
\text{target\_range}$ prevents RBF blindness in narrow target domains and applies
a priori without any experimental runs.
The \emph{stratified exploration rule} partitions the speed range into equal bands,
eliminating seed-sensitive failures and the max-speed local minima that trapped
random exploration at elevated heights.

\paragraph{Release-height and multi-arm generalisation.}
Applying the corrected algorithm with the two convergence rules achieves 100\% hit rates
across five release heights and three robot platforms, with mean landing errors below
35\,mm in all height configurations.
The decoupled architecture --- ball velocity set programmatically, arm kinematics
cosmetic --- provides portability without retraining.

\paragraph{Noise study.}
Theorem~\ref{thm:zeromean} proves that zero-mean symmetric noise is unlearnable:
a noise-aware policy achieves the same optimal mean action as a naive policy, while
correctly reporting higher expected cost.
Only biased noise (velocity slip, timing jitter, salt-and-pepper spikes) is learnable.
Experimental validation confirms 100\% hit rate for aware vs.\ 0\% for naive at 20\%
velocity slip.

\paragraph{Wind study.}
A complete and consistent wind pipeline (air-relative drag, 8-D$\to$11-D state
augmentation, updated GP and policy) was implemented and evaluated.
The central finding is counterintuitive: blind GP policies outperform explicit
wind-conditioned policies across all three wind regimes (constant, gusts, turbulence)
within the 15-trial budget.
The mechanism is the curse of dimensionality: the 4-D wind-aware policy space requires
$\approx 4\times$ more data than the 2-D blind policy space for equivalent coverage.
Blind GP treatment of wind as an implicit dynamics bias is strictly preferable until
the trial budget exceeds $\sim$40 trials.

\paragraph{Vision pipeline.}
Parallel OpenCV (HSV segmentation, depth back-projection, near-face bias correction)
and YOLOv8n (synthetic PyBullet dataset, bounding-box centroid back-projection)
pipelines were implemented and integrated with the trained policy for end-to-end
autonomous bin targeting.

\paragraph{Open questions.}
Future work includes hardware validation on physical manipulators; multi-seed
statistical evaluation of all five studies; wind-randomised training for a single
policy robust to varying airflow; and quantitative end-to-end evaluation of the vision
pipeline over a randomised target distribution.
