\section{Discussion}
\label{sec:discussion}

\subsection{Geometry-Dependence of Algorithm Defaults}
\label{sec:discussion:geometry}

The published Table~1 hyperparameters of \mcpilot{}~\cite{turcato2025mcpilot} are
calibrated to a Franka Panda arm on a flat floor with a fixed release height and a
target range of approximately 1.0\,m.
Our study quantifies how several of these defaults degrade silently when the geometry
changes.

\paragraph{RBF lengthscale.}
The default $\ell_s = 1.0$\,m is appropriate when the target arc spans $\sim$1\,m
in Euclidean distance.
At Config E ($z_{\mathrm{rel}} = 0.0$\,m, target range 0.35\,m), the same
lengthscale yields RBF sensitivity
$\exp\!\bigl(-\tfrac{1}{2}(0.35/1.0)^2\bigr) = 0.941$ between the two extremes of
the target domain---the policy produces an effectively constant output regardless of
target position.
Four successive failed experiments (Section~\ref{sec:study1}) were required to
isolate this as the root cause; once identified, replacing $\ell_s = 1.0$ with
$\ell_s = 0.15$\,m (maintaining sensitivity $\approx 0.066$, comparable to the
working configs) immediately recovered 10/10 hits.
The general rule is $\ell_s \approx 0.15 \times \text{target\_range}$
(Appendix~\ref{app:lengthscale}).

\paragraph{Cost lengthscale.}
The paper's $\ell_c = 0.1$\,m generates useful gradients only when the policy
already delivers within $\sim$0.15\,m of the target.
With random exploration, early landing errors span 0.3--1.5\,m, saturating the cost
at $c \approx 1.0$ and driving gradients to zero.
Widening to $\ell_c = 0.5$\,m keeps gradients alive across the full exploration
error range; then the cost can be tightened only after the GP becomes accurate
enough that near-misses are below 0.2\,m.
Attempting to revert to $\ell_c = 0.1$\,m after 5 trials still yields 0/10 hits
because early-trial errors remain too large.

\paragraph{Exploration coverage.}
At $z_{\mathrm{rel}} = 1.0$\,m, random exploration draws five speeds from
$[0, u_M]$; an unlucky seed (seed=1, Config B) draws four low speeds
($v < 1.5$\,m/s), leaving the GP uninformed about the high-speed dynamics needed
for far targets.
The policy converges to a max-speed local minimum (cost plateau at 0.70--0.73,
0/10 hits).
The identical seed succeeds with stratified exploration, which partitions
$[0, u_M]$ into $N_{\mathrm{exp}}$ equal bands and samples one throw per band,
guaranteeing full coverage in $N_{\mathrm{exp}}$ throws.

\subsection{Zero-Mean Noise as GP Regularisation}
\label{sec:discussion:noise}

An unexpected finding emerged from the noise study: the additive observation noise
$\sigma = 0.04$\,m/s (introduced to model velocity scatter) is not merely a noise
source for the aware/naive comparison---it is a necessary regulariser for GP
hyperparameter estimation.

When $\sigma = 0$ and exploration produces five noiseless, deterministic ballistic
trajectories, maximum likelihood estimation (MLE) of GP hyperparameters finds that
the mathematically easiest fit is a nearly flat function with effectively infinite
lengthscales (observed values: $\ell \approx 1791$\,m, $1602$\,m, $357$\,m over a
domain of $\sim$1\,m in position and $\sim$3\,m/s in velocity).
A GP with lengthscale $\gg$ domain width predicts the same landing position
regardless of launch speed, making policy gradients effectively zero.
With $\sigma = 0.04$, MLE must find finite lengthscales that explain why throws with
similar speeds land at different positions; the resulting structure is what allows the
aware policy to learn the $1/(1-\alpha)$ compensation multiplier
(Proposition~\ref{prop:zeromean} and Appendix~\ref{app:zeromean}).

\begin{proposition}[Zero-mean unlearnability]
\label{prop:zeromean}
Let $\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \Sigma_\epsilon)$ be
additive zero-mean noise on the release velocity.
The optimal policy $\pi^*_{\mathrm{aware}}$ minimising the noise-averaged cost
$J(\theta) = \mathbb{E}_{\boldsymbol{\epsilon}}\bigl[c(\mathbf{p}_T)\bigr]$ is
identical to the noiseless optimal policy $\pi^*_{\mathrm{naive}}$.
Consequently, noise awareness provides no improvement in aim under zero-mean noise.
\end{proposition}

\noindent The proof is given in Appendix~\ref{app:zeromean}.

The practical implication for simulation-only experiments is that practitioners
should add a small artificial observation noise $\sigma_n > 0$ to GP training data
even in noiseless simulators, treating it as GP regularisation rather than a physical
effect.
A value of $\sigma_n \approx 0.04$\,m/s is sufficient for the throwing domain.

\subsection{Decoupled Arm--RL Pipeline}
\label{sec:discussion:decoupled}

Setting the ball velocity explicitly via \texttt{resetBaseVelocity} at release
decouples arm tracking quality from GP training quality.
Even with substantial joint-tracking error (the iiwa7 position controller achieves
$\approx 7\%$ EE velocity tracking error in our experiments), the ball always
receives exactly $\mathbf{v}_{\mathrm{cmd}}$, and GP training data are never
corrupted by kinematic imprecision.

This design is not merely a simulation convenience.
It corresponds physically to a gripper that can be programmed to release at a
precisely measured EE velocity (read from the joint encoder forward kinematics).
In that case, the delivered velocity equals the measured velocity at the release
instant, and the arm controller only needs to bring the EE to the desired pose and
velocity---not to track a continuous trajectory with sub-millimetre precision.
The RL algorithm's data efficiency guarantees are preserved regardless of
trajectory-tracking quality, provided the release velocity is accurately measured or
commanded.

\subsection{Limitations}
\label{sec:discussion:limitations}

\paragraph{Simulation only.}
All experiments are conducted in simulation.
Hardware validation on a physical manipulator is needed to confirm that the
learned policies transfer and that the 5--10\% PyBullet--NumPy drag discrepancy
does not compound with real-world effects.

\paragraph{Fixed launch angle.}
The elevation angle $\alpha = 35^\circ$ is fixed across all experiments.
Joint optimisation of $\alpha$ and release speed would increase the reachable target
volume and potentially improve data efficiency by concentrating GP data along the
optimal ballistic arc.

\paragraph{Single-seed statistics.}
Most results are reported for seed=1 only; Study 1 Config B was also evaluated at
seed=2 for robustness verification.
Conference-level statistical claims require $\geq 5$ seeds to bound performance
variance.
The seed=1 vs seed=2 comparison for Config B (random exploration) illustrates that
hit rate can vary between 0/10 and 10/10 for an unlucky vs lucky seed, confirming
that single-seed results should be read as existence proofs rather than expected-case
performance.

\paragraph{Particle freeze approximation.}
Once a particle's $z$-coordinate falls below zero, its state is frozen at the
previous timestep rather than interpolated to the exact $z=0$ crossing.
This introduces a systematic over-estimation of the landing distance proportional
to the vertical velocity at the last step above ground.
For the simulation timestep $T_s = 0.02$\,s and typical vertical velocities of
1--3\,m/s at landing, the frozen-step error is 2--6\,cm per particle.
Landing detection in the actual rollout uses linear interpolation to correct this,
so the training data are accurate; the approximation affects only the Monte Carlo
cost estimate used for gradient computation (Appendix~\ref{app:freeze}).
