\section{Study 4: Wind Robustness}
\label{sec:study4}

\wip{Study 4 is planned but not yet executed.  This section describes the intended
experimental design; results will be added in the final version.}

Outdoor or atrium-deployed throwing systems must operate under non-zero ambient
airflow.
Study 4 extends the drag model (Eq.~\ref{eq:drag}) with a wind force term

\begin{equation}
  \mathbf{F}_W = \tfrac{1}{2}\rho\, C_D\, A\, \|\mathbf{w}\|\, \mathbf{w},
  \label{eq:wind}
\end{equation}

where $\mathbf{w} \in \mathbb{R}^3$ is the ambient wind velocity vector.
Three regimes are planned:

\begin{enumerate}
  \item \textbf{Constant wind.}
        A fixed $\mathbf{w} = (w_x, 0, 0)$ throughout all throws.
        The GP should absorb the systematic bias after a few trials, since
        constant wind shifts every landing by the same horizontal offset.

  \item \textbf{Gaussian gusts.}
        $\mathbf{w}$ is drawn independently at the start of each throw from
        $\mathcal{N}(\boldsymbol{\mu}_w, \Sigma_w)$.
        This is analogous to the velocity-slip noise studied in Study 2: the policy
        must compensate the mean wind while the variance sets an irreducible
        landing-error floor.

  \item \textbf{Per-timestep turbulence.}
        $\mathbf{w}$ is resampled at each integration step, approximating
        high-frequency turbulent fluctuations.
        The key question is whether the GP can model the resulting non-Gaussian
        landing distribution or whether wind state must be explicitly included in
        the policy input.
\end{enumerate}

The primary scientific question is whether augmenting the policy input with an
observed wind state $\mathbf{w}$ (\eg, from an anemometer) improves hit rate
compared with a wind-blind policy that relies on GP adaptation alone.
A secondary question is whether the zero-mean unlearnability result
(Section~\ref{sec:discussion:noise}) extends to zero-mean gust distributions, or
whether the momentum coupling through Eq.~\ref{eq:wind} makes even symmetric
gusts partially learnable.
