\section{System Architecture}
\label{sec:system}

\subsection{Two-Layer Decoupled Design}
\label{sec:system:design}

The implementation is organised around a strict two-layer separation that is the
central enabling design decision for portability across arms, noise conditions, and
wind regimes.

The \emph{algorithm layer} contains all learning components: a Gaussian-process
dynamics model, Monte Carlo policy gradient optimiser, radial-basis-function (RBF)
speed policy, and a ballistic simulator implemented in NumPy and PyTorch.

The \emph{simulation/visualisation layer} is either (a) the same NumPy ballistic
simulator used for training, or (b) a PyBullet scene with a manipulator arm (KUKA iiwa7,
Franka Panda, or xArm6 URDF) for validation and demonstration.

These two layers share a single interface: the commanded release velocity
$\mathbf{v}_{\mathrm{cmd}} \in \mathbb{R}^3$.
Ball velocity at release is set programmatically via PyBullet's
\texttt{resetBaseVelocity} to exactly $\mathbf{v}_{\mathrm{cmd}}$, or
$\mathbf{v}_{\mathrm{cmd}} + \boldsymbol{\epsilon}$ when noise is enabled.
Arm joint trajectories are planned and executed for visual fidelity, but any
joint-tracking error has no effect on the ball's initial velocity.
Consequently:
\begin{itemize}
  \item GP training data are never corrupted by kinematic imprecision.
  \item The algorithm's data-efficiency guarantees from \cite{turcato2025mcpilot}
        transfer intact across all arm URDFs (Study~2, Section~\ref{sec:study2}).
  \item Noise can be injected at the interface to model physical uncertainty
        independently of arm quality (Study~3, Section~\ref{sec:study3}).
  \item Wind augments the physics model without any change to the arm controller
        (Study~4, Section~\ref{sec:study4}).
\end{itemize}

\subsection{Sub-Project Structure}
\label{sec:system:subprojects}

The repository is organised into five sub-projects, each adding one independent
experimental variable:

\begin{itemize}
  \item \textbf{\texttt{mc-pilot/}} --- Baseline NumPy reproduction of \mcpilot{}
        \cite{turcato2025mcpilot}. Five targets on a flat plane ($z_{\mathrm{release}}
        = 0.5$\,m), no sensor noise. All four failure modes were diagnosed and fixed
        here (Section~\ref{sec:implementation}).

  \item \textbf{\texttt{mc-pilot-elevated/}} --- Release-height study. Sweeps five
        release heights $z \in \{0.0, 0.5, 1.0, 1.5, 2.0\}$\,m using the NumPy
        ballistic simulator. Isolates the effect of geometric variation on GP
        generalisation.

  \item \textbf{\texttt{mc-pilot-pybullet/}} --- PyBullet arm integration. Introduces
        the KUKA iiwa7 visualisation pipeline and injected noise models for
        Study~3 (Section~\ref{sec:study3}).

  \item \textbf{\texttt{mc-pilot-pb-elevated/}} --- Combined PyBullet scene with
        elevated release positions and multi-arm URDF support.
        Used for Study~2 (Section~\ref{sec:study2}) and end-to-end visualisation.

  \item \textbf{\texttt{mc-pilot-wind/}} --- Wind extension. Adds air-relative drag,
        11-D state augmentation, wind-aware GP and policy, and three wind models for
        Study~4 (Section~\ref{sec:study4}).
\end{itemize}

\subsection{NumPy Ballistic Simulator}
\label{sec:system:numpy}

The NumPy ballistic simulator integrates the 6D ball state
$\mathbf{s} = [x, y, z, v_x, v_y, v_z]^\top$ under gravity and aerodynamic drag.
The state update follows the reconstruction rule (Eqs.~\ref{eq:vel_update}--\ref{eq:pos_update}).
The 6D state is augmented to 8D by appending target coordinates $[P_x, P_y]$ to form
the GP input.
Landing detection at $z \leq 0$ uses linear interpolation between the last airborne and
first underground states to recover the exact ground-crossing position.

\subsection{PyBullet Arm Pipeline}
\label{sec:system:pybullet}

The \texttt{ArmController} class wraps a manipulator URDF loaded from
\texttt{pybullet\_data}.
Throw execution proceeds in five steps:

\begin{enumerate}
  \item \textbf{IK for release pose.}  Desired release position $\to$ joint angles
        via PyBullet's built-in IK solver (200 iterations, threshold $10^{-4}$).

  \item \textbf{Joint velocity mapping.}  Jacobian pseudoinverse maps desired
        end-effector velocity to joint velocities, scaled down if any joint exceeds
        its velocity limit.

  \item \textbf{Three-phase cubic trajectory.}  A piecewise cubic spline drives
        neutral $\to$ windup $\to$ release $\to$ follow-through.

  \item \textbf{Ball attachment and release.}  A \texttt{JOINT\_FIXED} constraint
        couples ball to end-effector; removed at release timestep.

  \item \textbf{Velocity override.}  \texttt{resetBaseVelocity} sets ball velocity to
        $\mathbf{v}_{\mathrm{cmd}}$. Post-release drag applied manually via
        \texttt{applyExternalForce}.
\end{enumerate}

The \texttt{vel\_limit\_multiplier} parameter governs whether a feasibility warning
is issued; it does not alter the ball's trajectory.
For elevated release heights (Study~1: $z_{\mathrm{rel}} = 1.0$--$2.0$\,m), the arm
base is elevated on a pedestal: \texttt{base\_position} $= (0, 0, z_{\mathrm{rel}} - 0.5)$.
This places the neutral EE height at $z_{\mathrm{rel}} + 0.21$\,m, keeping the IK
target within the arm's kinematic range.

\subsection{PyBullet--NumPy Physics Discrepancy}
\label{sec:system:discrepancy}

Drag applied via \texttt{applyExternalForce} per simulation step accumulates small
numerical differences compared to the closed-form NumPy midpoint rule.
At $u_M = 3.5$\,m/s and $z_{\mathrm{rel}} = 1.0$\,m, the landing-range discrepancy
is $\approx 6\%$.
This does not impair learning: the GP observes PyBullet ground truth during both
exploration and policy trials, so both the dynamics model and policy converge to the
PyBullet physics, not the NumPy approximation.
